# 2.11.2020 

# **Analysis, Exercise 2**


**Description of analysed data**

The aim of this study was to set light on the relationship between learning approaches and students' achievements in an introductory statistics course in Finland.

The data set (N = 166) used in this report consisted of 7 variables: Gender, Age, attitude, Deep approach, Surface approach, Strategic approach and total points. The participants were 17-55-years-olds (mean = 25.5) and 66 % of them were Female.

Read more about the used scales and variables from 
[here](https://www.mv.helsinki.fi/home/kvehkala/JYTmooc/JYTOPKYS2-meta.txt)

R script for graphical overview of the data:

```{r}
library()
analysisdata <- read.table("http://s3.amazonaws.com/assets.datacamp.com/production/course_2218/datasets/learning2014.txt", sep=",", header=TRUE)

#Description of data set
library(psych)

describe(analysisdata[-1])
```
***

##### **Here's a little glossary to help understand the short terms used in tables:**
**deep** =  Deep approach

**surf** = Surface approach

**stra** = Strategic approach

**points** = total points

**F** = Female

**M** = Male


See the means, standard deviations, medians, minimum and maxim values from the table above. The table also shows skewness and kuriosis which tell us the normality of the variables. All the sum variables were normally distributed. 


##### **Bellow you can find information about gender:**

***
```{r}

table(analysisdata$gender)

```

#### **A Plot Matrix**

Find bellow a plot matrix that includes all the variables. As you can see from the matrix, there are great significant correlation between surface approach & deep approach and attitude & points.

```{r}

# creating a plot matrix out of all the variables.
library(GGally)
library(ggplot2)

p <- ggpairs(analysisdata, mapping = aes(alpha = 0.3), lower = list(combo = wrap("facethist", bins = 20)))

p

```

#### Creating a regression model with explanatory variables

I chose surf, deep and attitude as the predicted variables for this exercise. Total points were my dependent variable.

Bellow you can find model fit indexes.

```{r}
# creating a regression model with multiple explanatory variables
my_model <- lm(points ~ attitude + deep + surf, data = analysisdata)

# printing out a summary of the model
summary(my_model)

```
#### **Interpretation of the model fit**

The indexes above show that the model in not working as it is. F-statistic's p-value is greater than 0.05 which means that the overall model is not significant. The only working predicted variable seems to be attitude.

Deep and surf didn't have significant relationship with the dependent variable. They were removed from the model bellow. Since there is only one predicted variable now, the overall model fit is still not significant.

```{r}
# creating a regression model
my_model2 <- lm(points ~ attitude, data = analysisdata)

# printing out a summary of the model
summary(my_model2)

```

#### **Results of the regression analysis**

Seems like only attitude predicts overall exam points. Good attitude predicts high exam points.The multiple of R squared of the model shows that 19% of the variation is points is explained by the attitude.

#### **Residuals vs Fitted values, Normal QQ-plot and Residuals vs Leverage**

As you can see bellow *Residual vs. Fitter points* spread randomly. *Normal Q-Q plot points* fits quite well to the line which indicates that the model is normally distributed. There are no outliners standing out in *Residuals vs. Leverage plot*.

```{r}
par(mfrow = c(2,2))
plot(my_model2, which = c(1, 2, 5))

```
