# 2.11.2020 

# **Analysis, Exercise 2**


**Description of analysed data**

The aim of this study was to set light on the relationship between learning approaches and students' achievements in an introductory statistics course in Finland.

The data set (N = 166) used in this report consisted of 7 variables: Gender, Age, attitude, Deep approach, Surface approach, Strategic approach and total points. The participants were 17-55-years-olds (mean = 25.5) and 66 % of them were Female.

Read more about the used scales and variables from 
[here](https://www.mv.helsinki.fi/home/kvehkala/JYTmooc/JYTOPKYS2-meta.txt)

R script for graphical overview of the data:

```{r}
library()
analysisdata <- read.table("http://s3.amazonaws.com/assets.datacamp.com/production/course_2218/datasets/learning2014.txt", sep=",", header=TRUE)

#Description of data set
library(psych)

describe(analysisdata[-1])
```
***

##### **Here's a little glossary to help understand the short terms used in tables:**
**deep** =  Deep approach

**surf** = Surface approach

**stra** = Strategic approach

**points** = total points

**F** = Female

**M** = Male


See the means, standard deviations, medians, minimum and maxim values from the table above. The table also shows skewness and kuriosis which tell us the normality of the variables. All the sum variables were normally distributed. 


##### **Bellow you can find information about gender:**

***
```{r}

table(analysisdata$gender)

```

#### **A Plot Matrix**

Find bellow a plot matrix that includes all the variables refined based on gender. As you can see from the matrix, there are great significant correlation between surface approach & deep approach and attitude & points.

```{r}
# creating a plot matrix out of all the variables to see how females and males differs from each other.
library(GGally)
library(ggplot2)

p <- ggpairs(analysisdata, mapping = aes(col = gender, alpha = 0.3), lower = list(combo = wrap("facethist", bins = 20)))

p
```

#### Creating a regression model with explanatory variables

I chose surf, deep and attitude variables for this exercise. Total points were my dependent variable.

Bellow you can find model fit indexes.

```{r}
# creating a regression model with multiple explanatory variables
my_model <- lm(points ~ attitude + deep + surf, data = analysisdata)

# printing out a summary of the model
summary(my_model)

```

Choose three variables as explanatory variables and fit a regression model where exam points is the target (dependent) variable. Show a summary of the fitted model and comment and interpret the results. Explain and interpret the statistical test related to the model parameters. If an explanatory variable in your model does not have a statistically significant relationship with the target variable, remove the variable from the model and fit the model again without it. (0-4 points)



Using a summary of your fitted model, explain the relationship between the chosen explanatory variables and the target variable (interpret the model parameters). Explain and interpret the multiple R squared of the model. (0-3 points)

Produce the following diagnostic plots: Residuals vs Fitted values, Normal QQ-plot and Residuals vs Leverage. Explain the assumptions of the model and interpret the validity of those assumptions based on the diagnostic plots. (0-3 points)